Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-23 analysis/paper/paper.Rmd
finished the background section

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail and broomcorn millet based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca. BP [@Hao2022]. It is suggested that mixed farming of millet and rice could spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago based on the evidence of rice remains from Guijiabao and other sites in the neighboring areas, Baiyangcun and Henglanshan [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. Gaoshan could be even older than the Baodun culture according to the burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].

A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This finding also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice and millet cultivation with domesticated pigs.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].

@@

# Discussion

Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals. Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot. A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.

>>>>

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca. BP [@Hao2022].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].

A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.

It is generally accepted that mixed farming of millet and rice spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010]. The current finding at Guijiabao and Gaoshan support this argument. However, the ratio of rice to millet grains found at Guijiabao suggests that millet is the major crop in an earlier period, which is different from a clear predominance of rice (*Oryza sativa*) at Gaoshan or other sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study provides direct evidence from food residue in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and another sample was from the 2016 season. To understand the content and use of pottery vessels in Yanyuan in an earlier period, we analyzed potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].

@@

# Discussion
Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals. Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot. A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.
